\batcave: Bayesian Analysis Tools for Context-Aware Variant Evaluation

### 2019-09-18 NAR_DRAFT.tex
added second signature figure and fixed several things in KL figure

\begin{figure}
  \begin{center}
  \includegraphics{figures/signatures_only.pdf}
  \end{center}
  \bkmcomment{tumor in F is probably normal breast plus APOBEC. Mention somewhere?}
  \caption{Tumor mutation profiles.
  In each panel, the x-axis corresponds to each of the 96 possible substitution types, and the y-axis is the proportion of total mutations of each type.
  \textbf{(A)} The observed mutation profile of an acute myeloid leukemia used in our real data analysis \cite{Griffith2015}.
  \textbf{(B)} The observed mutation profile of a breast tumor used in our real data analysis \cite{Shi2018}.
  \textbf{(C)}\&\textbf{(D)} The observed mutation profiles of two breast tumors \cite{Alexandrov2019}.
  \textbf{()} A mutation profile used for simulating tumors, made up of equal proportions of COSMIC mutation signatures 1, 7, \& 11.
  \textbf{()} Equal proportions of signatures 1, 4, \& 5.
  \textbf{()} Equal proportions of signatures 1, 3, \& 5.
  \rngcomment{Move simulated profiles to own figure.}
   }
  \label{NAR-sigfig}
 \end{figure}

@@

\subsection{Tumor simulations}
We used a neutral branching process with no death and $\mu=3\cdot10^{-6}$ to simulate realistic distributions of allele frequencies.
Tumors were simulated with three different mutation profiles composed of COSMIC mutation signatures (version 2) \cite{XXX}.
\rngcomment{Put COSMIC website as a citation.}
Each simulated profile includes COSMIC signature 1, which is found in nearly all tumors and is associated with spontaneous cytosine deamination.
The ``Concentrated'' profile (Fig.~\ref{XXX}) is an equal combination of COSMIC signatures 1, 7, and 11, which has a large percentage of C $>$ T substitutions such as are often seen in cancers caused by UV exposure.
The ``Intermediate'' profile (Fig.~\ref{XXX}) is an equal combination of COSMIC signatures 1, 4, and 5, which has been associated with tobacco carcinogens and is representative of some lung cancers.
The ``Diffuse'' profile (Fig.~\ref{XXX}) is an equal combination of COSMIC signatures 1, 3, and 5, which has been associated with inactivating germline mutations in the BRCA1/2 genes leading to a deficiency in DNA double strand break repair. \rngcomment{Citations for these signature combinations.}
\bkmcomment{These are really from the text of the cosmic website, the citation is probably Alexandrov2015 or Alexandrov2013a}
Simulated variants were sampled from a combination of the Cancer Genome Atlas (TCGA) and Pan-Cancer Analysis of Whole Genomes (PCAWG) databases, which include mutations found in all types of cancer.
Whole genome (100X depth) and whole exome (500X depth) reads were simulated from the GRCh38 reference genome using VarSim \cite{Mu2015} and aligned with BWA \cite{Li2009a}, both with default parameters.
Variants were inserted to create tumors with BAMSurgeon with default parameters~\citep{Ewing2015a} and called with MuTect 1.1.7~\cite{Cibulskis2013} with the following parameters:
\begin{scriptsize}
\begin{verbatim}
java -Xmx24g -jar $MUTECT_JAR --analysis_type MuTect \
        --reference_sequence $ref_path \
        --dbsnp $db_snp \
        --enable_extended_output \
        --fraction_contamination 0.00 \
        --tumor_f_pretest 0.00 \
        --initial_tumor_lod -10.00 \
        --required_maximum_alt_allele_mapping_quality_score 1 \
        --input_file:normal $tmp_normal \
        --input_file:tumor $tmp_tumor \
        --out $out_path/$chr.txt \
        --coverage_file $out_path/$chr.cov
\end{verbatim}
\end{scriptsize}
Variants identified by MuTect are labelled as to whether they pass all filters, fail to pass only the the evidence threshold \textrm{tlod\_f\_star} filter, or fail to pass any other filter. 
Variants that passed all filters or failed only \textrm{tlod\_f\_star} were then passed to \batcave for prior estimation and rescoring.

@@

\begin{figure}
  \begin{center}
  \includegraphics{figures/kl_only.pdf}
  \end{center}
  \caption{Convergence of the site-specific prior to the data generating distribution. Plotted is the Kullback-Leibler divergence between the estimated and simulated profiles versus number of incorporated mutations, for \textbf{(A)} whole exomes and \textbf{(B)} whole genomes.
  \rngcomment{Zoom panel B into relevant region of plot ($\sim$1000 mutations). What is notation (target || prior)?}}\bkmcomment{just saw it somewhere, change to comma.}
\label{NAR-kl_fig}
\end{figure}

>>>>

\begin{figure}
  \begin{center}
  \includegraphics{figures/real_signatures_only.pdf}
  \end{center}
  \caption{Tumor mutation profiles.
  In each panel, the x-axis corresponds to each of the 96 possible substitution types, and the y-axis is the proportion of total mutations of each type.
  \textbf{(A)} The observed mutation profile of an acute myeloid leukemia used in our real data analysis \cite{Griffith2015}.
  \textbf{(B)} The observed mutation profile of a breast tumor used in our real data analysis \cite{Shi2018}.
  \textbf{(C)}\&\textbf{(D)} The observed mutation profiles of two breast tumors \cite{Alexandrov2019}.
   }
  \label{NAR-realsigfig}
 \end{figure}

@@

\begin{figure}
  \begin{center}
  \includegraphics{figures/sim_signatures_only.pdf}
  \end{center}
  \bkmcomment{tumor in F is probably normal breast plus APOBEC. Mention somewhere?}
  \caption{Tumor mutation profiles.
  In each panel, the x-axis corresponds to each of the 96 possible substitution types, and the y-axis is the proportion of total mutations of each type.
  \textbf{(A)} A mutation profile used for simulating tumors, made up of equal proportions of COSMIC mutation signatures 1, 7, \& 11.
  \textbf{(B)} Equal proportions of signatures 1, 4, \& 5.
  \textbf{(C)} Equal proportions of signatures 1, 3, \& 5.
   }
  \label{NAR-simsigfig}
 \end{figure}
\subsection{Tumor simulations}
We used a neutral branching process with no death and $\mu=3\cdot10^{-6}$ to simulate realistic distributions of allele frequencies.
Tumors were simulated with three different mutation profiles composed of COSMIC mutation signatures (version 2) \cite{XXX}.
\rngcomment{Put COSMIC website as a citation.}
Each simulated profile includes COSMIC signature 1, which is found in nearly all tumors and is associated with spontaneous cytosine deamination.
The ``Concentrated'' profile (Fig.~\ref{simsigfig}\textbf{A}) is an equal combination of COSMIC signatures 1, 7, and 11, which has a large percentage of C $>$ T substitutions such as are often seen in cancers caused by UV exposure.
The ``Intermediate'' profile (Fig.~\ref{simsigfig}\textbf{B}) is an equal combination of COSMIC signatures 1, 4, and 5, which has been associated with tobacco carcinogens and is representative of some lung cancers.
The ``Diffuse'' profile (Fig.~\ref{simsigfig}\textbf{C}) is an equal combination of COSMIC signatures 1, 3, and 5, which has been associated with inactivating germline mutations in the BRCA1/2 genes leading to a deficiency in DNA double strand break repair. \rngcomment{Citations for these signature combinations.}
\bkmcomment{These are really from the text of the cosmic website, the citation is probably Alexandrov2015 or Alexandrov2013a}
Simulated variants were sampled from a combination of the Cancer Genome Atlas (TCGA) and Pan-Cancer Analysis of Whole Genomes (PCAWG) databases, which include mutations found in all types of cancer.
Whole genome (100X depth) and whole exome (500X depth) reads were simulated from the GRCh38 reference genome using VarSim \cite{Mu2015} and aligned with BWA \cite{Li2009a}, both with default parameters.
Variants were inserted to create tumors with BAMSurgeon with default parameters~\citep{Ewing2015a} and called with MuTect 1.1.7~\cite{Cibulskis2013} with the following parameters:
\begin{scriptsize}
\begin{verbatim}
java -Xmx24g -jar $MUTECT_JAR --analysis_type MuTect \
        --reference_sequence $ref_path \
        --dbsnp $db_snp \
        --enable_extended_output \
        --fraction_contamination 0.00 \
        --tumor_f_pretest 0.00 \
        --initial_tumor_lod -10.00 \
        --required_maximum_alt_allele_mapping_quality_score 1 \
        --input_file:normal $tmp_normal \
        --input_file:tumor $tmp_tumor \
        --out $out_path/$chr.txt \
        --coverage_file $out_path/$chr.cov
\end{verbatim}
\end{scriptsize}
Variants identified by MuTect are labelled as to whether they pass all filters, fail to pass only the the evidence threshold \textrm{tlod\_f\_star} filter, or fail to pass any other filter. 
Variants that passed all filters or failed only \textrm{tlod\_f\_star} were then passed to \batcave for prior estimation and rescoring.

@@

\begin{figure}
  \begin{center}
  \includegraphics{figures/kl_only.pdf}
  \end{center}
  \caption{Convergence of the site-specific prior to the data generating distribution. Plotted is the Kullback-Leibler divergence between the estimated and simulated profiles versus number of incorporated mutations, for \textbf{(A)} whole exomes and \textbf{(B)} whole genomes.
  }
\label{NAR-kl_fig}
\end{figure}
